---
title: "Numbat haplotype-aware HMM"
description: "Detecting CNVs using Numbat haplotype-aware HMM from bulk and single-cell RNA-seq data"
documentclass: book
output:
  bookdown::gitbook:
    split_by: rmd
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

The Numbat haplotype-aware HMM leverages [population-based phasing](https://en.wikipedia.org/wiki/Haplotype_estimation) to sensitively detect CNVs from gene expression data. It can be used to analyze scRNA-seq (after aggregating cells into "pseudobulks") as well as bulk RNA-seq data. In this tutorial, we will walk through some examples of both use cases.

```{r}
library(numbat)
library(dplyr)
library(glue)
library(stringr)
library(data.table)
```

# Analyzing scRNA-seq data
```{r}

```

# Analyzing bulk RNA-seq data
The Numbat HMM can also be used to detect CNVs from bulk RNA-seq data. 
```{r}
allele_counts = fread('http://pklab.med.harvard.edu/teng/data/bulk_example/MN-5_TUMOR_allele_counts.tsv.gz')
gene_counts = readRDS(url('http://pklab.med.harvard.edu/teng/data/bulk_example/gene_counts.rds'))
```
Sample MN-1037 has a diploid genome so we can use it to create a reference expression profile.
```{r}
ref_internal = gene_counts[,'MN-1037_TUMOR',drop=F] %>% {./sum(.)}
head(ref_internal)
```
We can now analyze samples using the Numbat HMM.
```{r, fig.width = 12, fig.height=2, dpi=300}
sample = 'MN-5_TUMOR'

bulk = get_bulk(
        count_mat = gene_counts[,sample,drop=F],
        df_allele = allele_counts,
        lambdas_ref = ref_internal,
        genetic_map = genetic_map_hg38,
        gtf = gtf_hg38
    ) %>% 
    analyze_bulk()

bulk %>% plot_psbulk(min_depth = 15)
```

